---
title: "MID/CMB Strep group distributions"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---

```{r load_libraries, echo = F, message = F}
library(knitr)
library(data.table)
library(janitor)
library(tidyverse)
library(cowplot)
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r strep_groups_incomplete, eval = F}
# read in the Strep group table James made for the Deep Evo project

strep_groups <- fread("./00-documentation.backup/25-streptococcus_cladegroup_amylasegroup_database.tsv") %>%
  select(Taxon, dRep) %>%
  rename(Species = Taxon) %>%
  mutate(Species = str_replace_all(Species, "_", " "),
         dRep = str_replace_all(dRep, "NA","Unknown"),
         dRep = str_replace_all(dRep, "sanguinis","Sanguinis"),
         dRep = str_replace_all(dRep, "mitis","Mitis"),
         dRep = str_replace_all(dRep, "salivarius","Salivarius"),
         dRep = str_replace_all(dRep, "anginosus","Anginosus"),
         dRep = str_replace_all(dRep, "bovis","Bovis"),
         dRep = str_replace_all(dRep, "pyogenic","Pyogenic"),
         dRep = str_replace_all(dRep, "mutans","Mutans")) %>%
  mutate(dRep = fct_relevel(dRep, "Sanguinis","Mitis","Salivarius","Anginosus","Bovis","Pyogenic","Mutans","Unknown"))

```

```{r}

strep_groups <- fread("./00-documentation.backup/dRep_strep_groups.tsv") %>%
  # remove the extrac column with the old Clade_consensus
  select(-Clade_Consensus) %>%
  # make a factor for plotting
  mutate(dRep = fct_relevel(dRep, "Sanguinis","Mitis","Salivarius","Anginosus","Bovis","Pyogenic","Mutans","Other_oral","Other","Unknown"))

```


```{r metadata}

load("./05-results.backup/Metadata_tables.RData")

```


```{r}
# set colors to resemble James' figure

clade_colors <- c("Sanguinis" = "#a50026",
                  "Mitis" = "#d73027",
                  "Salivarius" = "#f46d43",
                  "Anginosus" = "#fee090",
                  "Bovis" = "#abd9e9",
                  "Pyogenic" = "#74add1",
                  "Mutans" = "#4575b4",
                  "Oral_other" = "#313695",
                  "Other" = "#4d4d4d",
                  "Unknown" = "#e0e0e0")

genus_colors <- c("Streptococcus" = "#a50026", "Other" = "#e0e0e0")

```


## MALT RefSeq
```{r load_MALT_RefSeq_data}

# load the MALT RefSeq taxonomy tables that have been decontaminated and the column headers are cleaned up
load("./05-results.backup/Taxonomy_tables.RData")

# list poor samples from cuperdec
poor_samples_rs <- fread("./05-results.backup/cuperdec_poor_samples.tsv") %>%
  rename(Library_ID = Sample) %>%
  filter(str_detect(Library_ID, "MID")) %>%
  pull(Library_ID)

# we'll use the prevalence-filtered dataset
malt_rs <- all_species_decontam_prev %>%
  select(matches("Species|MID|CMB")) %>%
  adorn_totals(where = "col")  %>%
  filter(Total >  0) %>%
  select(-Total) %>%
  select(-poor_samples_rs)


```

First do a PCA to get the order of samples in the horseshoe
```{r pca_rs}
# prep the table

outliers <- c("EXB059.A2501","LIB058.A0103","LIB058.A0106","CSS","CSD")
poor_samples <- fread("./05-results.backup/cuperdec_poor_samples.tsv") %>%
  pull(Sample)

malt_otu_pca_table <- malt_rs %>%
  select(-one_of(outliers)) %>%
  gather("Library_ID","Counts",2:ncol(.)) %>%
  mutate(Counts = Counts + 1) %>%
  spread(Species, Counts) %>%
  column_to_rownames("Library_ID")

# perform a PCA to see how the data cluster
malt_otu.pca <- mixOmics::pca(malt_otu_pca_table, ncomp = 2, logratio = 'CLR')
malt_pca_values <- malt_otu.pca$variates$X %>%
  as.data.frame() %>%
  rownames_to_column("Library_ID") %>%
  inner_join(., full_metadata, by = "Library_ID")

mid_pc1_sample_order <- malt_pca_values %>%
  arrange(PC1) %>%
  pull(Library_ID)

ggplot(malt_pca_values, aes(PC1, PC2, color = Region, shape = Region)) +
  geom_text(aes(label = Library_ID), size = 3) +
  theme_minimal()

```


```{r malt_rs_graphs}

# set the order of samples based on proportion of Sanguinis and agninosis
# need to get the proportion of each for this
# then arrange by proportion

sample_order_rs <- malt_rs  %>%
  filter(str_detect(Species, "Streptococcus")) %>%
  # join with the Strep grup table for plotting
  inner_join(., strep_groups) %>%
  select(-Species) %>%
  select(dRep, everything()) %>%
  # long-format
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  # sum together all species per group
  group_by(Library_ID, dRep) %>%
  summarize(Reads = sum(Counts)) %>%
  # transpose
  spread(dRep, Reads) %>%
  # convert table to proportions
  adorn_percentages(denominator = "row") %>%
  # arrange by Sanguinis and for the ones w/o Sanguinis, by Anginosus
  arrange(desc(Sanguinis), Anginosus)%>%
  pull(Library_ID)


malt_rs %>%
  filter(str_detect(Species, "Streptococcus")) %>%
  # join with the Strep grup table for plotting
  inner_join(., strep_groups) %>%
  select(Species, dRep, everything()) %>%
  # convert to long-format
  gather("Library_ID","Counts", 3:ncol(.)) %>%
  mutate(Library_ID = fct_relevel(Library_ID, sample_order_rs)) %>%
  dplyr::group_by(Library_ID, dRep) %>%
  summarize(Count_sum = sum(Counts)) %>%
  ggplot(., aes(x = Library_ID, y = Count_sum, fill = dRep)) +
         geom_bar(position = "fill", stat = "identity") + # , color = "black"
         theme_minimal(base_size = 14) +
         scale_fill_manual(values = clade_colors) +
         theme(axis.text.x = element_text(angle = 90, hjust = 1),
               axis.text = element_text(size = 8)) +
         # scale_y_continuous(breaks=scales::pretty_breaks(n=6)) +
         geom_hline(yintercept = 0.5, linetype = "dotted") +
         ylab("Proportion Strep group") +
         ggtitle("")

```

```{r save_rs_order, eval = F}

sample_order_rs %>%
  as.data.frame(.) %>%
 rename(Library_ID = ".") %>%
fwrite(., "./05-results.backup/malt_rs_Strep_group_order.tsv", sep = "\t", quote = F)

```

```{r}

#plot ordered by horseshoe
malt_rs %>%
  filter(str_detect(Species, "Streptococcus")) %>%
  # join with the Strep grup table for plotting
  inner_join(., strep_groups) %>%
  select(Species, dRep, everything()) %>%
  # convert to long-format
  gather("Library_ID","Counts", 3:ncol(.)) %>%
  mutate(Library_ID = fct_relevel(Library_ID, mid_pc1_sample_order)) %>%
  mutate(Library_ID = fct_rev(Library_ID)) %>%
  dplyr::group_by(Library_ID, dRep) %>%
  summarize(Count_sum = sum(Counts)) %>%
  ggplot(., aes(x = Library_ID, y = Count_sum, fill = dRep)) +
         geom_bar(position = "fill", stat = "identity") + # , color = "black"
         theme_minimal(base_size = 14) +
         scale_fill_manual(values = clade_colors) +
         theme(axis.text.x = element_text(angle = 90, hjust = 1),
               axis.text = element_text(size = 8)) +
         # scale_y_continuous(breaks=scales::pretty_breaks(n=6)) +
         ylab("Proportion Strep group") +
         ggtitle("")


```


```{r rs_strep_proportion}

strep_prop <- malt_rs %>%
  # make a column with Strep or other
  mutate(Genus = ifelse(str_detect(Species,"Strep"),"Streptococcus","Other")) %>%
  # remove Species column leaving Genus column
  select(Genus, everything()) %>%
  select(-Species) %>%
  # convert to long format
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  group_by(Library_ID, Genus) %>%
  # sum all Strep read counts and all Other read counts
  summarize(Reads = sum(Counts)) %>%
  spread(Genus, Reads) %>%
  # convert to proportions
  adorn_percentages() %>%
  ungroup()

strep_prop %>%
  gather("Genus","Counts", 2:ncol(.)) %>%
  mutate(Library_ID = fct_relevel(Library_ID, sample_order_rs)) %>%
  arrange(Library_ID) %>%
  ggplot(., aes(x = Library_ID, y = Counts, fill = Genus)) +
         geom_bar(stat = "identity") + # , color = "black"
         theme_minimal(base_size = 14) +
         scale_fill_manual(values = genus_colors) +
         theme(axis.text.x = element_text(angle = 90, hjust = 1),
               axis.text = element_text(size = 8)) +
         # scale_y_continuous(breaks=scales::pretty_breaks(n=6)) +
         ylab("Proportion Strep group") +
         ggtitle("")

```


## MALT nt
```{r load_MALT_nt_data}
# read in the species table
nt_raw <- fread("./05-results.backup/MID_CMB_malt_nt-comparison-species.txt")

# clean the file names
nt_raw <- rename(nt_raw, Species = `#Datasets`)
colnames(nt_raw) <- gsub(".unmapped","", colnames(nt_raw))
colnames(nt_raw) <- gsub(".10M","", colnames(nt_raw))

# remove human and unassigned reads, convert NA to 0, and remove the 8 MID samples with no metadata
nt_raw <- nt_raw %>% 
  filter(!str_detect(Species, "Homo sapiens|Not assigned")) %>%
  replace(is.na(.), 0) %>%
  # these samples have no metadata
  select(-c("MID025.A0101","MID033.A0101","MID052.A0101","MID056.A0101","MID065.A0101","MID068.A0101","MID076.A0101","MID078.A0101")) %>%
  arrange(Species)

```

Now decontam the tables
```{r nt_decontam}
nt_contaminants <- fread("./05-results.backup/contaminant_species_nt.tsv", sep = "\t")
poor_samples_nt <- fread("./05-results.backup/cuperdec_nt_poor_samples.tsv") %>%
  pull(Sample)

# remove contaminant species from nt table
nt_species_decontam <- nt_raw %>%
  anti_join(., nt_contaminants) %>%
  select(-poor_samples_nt)

# and remove poorly preserved samples
malt_nt <- nt_species_decontam %>%
  select(matches("Species|MID|CMB")) %>%
  adorn_totals(where = "col")  %>%
  filter(Total >  0) %>%
  select(-Total)

```


```{r pca_nt}

# prep the table

malt_otu_pca_table_nt <- malt_nt %>%
  select(-one_of(outliers)) %>%
  gather("Library_ID","Counts",2:ncol(.)) %>%
  mutate(Counts = Counts + 1) %>%
  spread(Species, Counts) %>%
  column_to_rownames("Library_ID")

# perform a PCA to see how the data cluster
malt_otu_nt.pca <- mixOmics::pca(malt_otu_pca_table_nt, ncomp = 2, logratio = 'CLR')
malt_pca_values_nt <- malt_otu_nt.pca$variates$X %>%
  as.data.frame() %>%
  rownames_to_column("Library_ID") %>%
  inner_join(., full_metadata, by = "Library_ID")

mid_pc1_sample_order_nt <- malt_pca_values_nt %>%
  arrange(PC1) %>%
  pull(Library_ID)

ggplot(malt_pca_values_nt, aes(PC1, PC2, color = Region, shape = Region)) +
  geom_text(aes(label = Library_ID), size = 3) +
  theme_minimal()


```




```{r}

sample_order_nt <- malt_nt %>%
  filter(str_detect(Species, "Streptococcus")) %>%
  # join with the Strep grup table for plotting
  inner_join(., strep_groups) %>%
  select(-Species) %>%
  select(dRep, everything()) %>%
  # transpose
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  group_by(Library_ID, dRep) %>%
  summarize(Reads = sum(Counts)) %>%
  spread(dRep, Reads) %>%
  # convert table to proportions
  adorn_percentages(denominator = "row") %>%
  # arrange by Sanguinis and for the ones w/o Sanguinis, by anginosis
  arrange(desc(Sanguinis), Anginosus) %>%
  pull(Library_ID)


plot_nt <- malt_nt %>%
  filter(str_detect(Species, "Streptococcus")) %>%
  # join with the Strep grup table for plotting
  inner_join(., strep_groups) %>%
  select(Species, dRep, everything()) %>%
  # convert to long-format
  gather("Library_ID","Counts", 3:ncol(.)) %>%
  mutate(Library_ID = fct_relevel(Library_ID, sample_order_nt)) %>%
  dplyr::group_by(Library_ID, dRep) %>%
  summarize(Count_sum = sum(Counts)) %>%
  ggplot(., aes(x = Library_ID, y = Count_sum, fill = dRep)) +
         geom_bar(position = "fill", stat = "identity") + # , color = "black"
         theme_minimal(base_size = 14) +
         scale_fill_manual(values = clade_colors) +
         theme(axis.text.x = element_text(angle = 90, hjust = 1),
               axis.text = element_text(size = 8)) +
         # scale_y_continuous(breaks=scales::pretty_breaks(n=6)) +
         ylab("Proportion Strep group") +
         ggtitle("")

plot_nt

# ggsave("~/Desktop/MID_strep_groups_MALT_nt.pdf", plot = plot_nt, device = "pdf",
#        scale = 1, width = 9, height = 5, units = c("in"), dpi = 300)

```

```{r save_nt_order, eval = F}

sample_order_nt %>%
  as.data.frame(.) %>%
 rename(Library_ID = ".") %>%
fwrite(., "./05-results.backup/malt_nt_Strep_group_order.tsv", sep = "\t", quote = F)

```

```{r}
#plot ordered by horseshoe
malt_nt %>%
  filter(str_detect(Species, "Streptococcus")) %>%
  # join with the Strep grup table for plotting
  inner_join(., strep_groups) %>%
  select(Species, dRep, everything()) %>%
  # convert to long-format
  gather("Library_ID","Counts", 3:ncol(.)) %>%
  mutate(Library_ID = fct_relevel(Library_ID, mid_pc1_sample_order_nt)) %>%
  mutate(Library_ID = fct_rev(Library_ID)) %>%
  dplyr::group_by(Library_ID, dRep) %>%
  summarize(Count_sum = sum(Counts)) %>%
  ggplot(., aes(x = Library_ID, y = Count_sum, fill = dRep)) +
         geom_bar(position = "fill", stat = "identity") + # , color = "black"
         theme_minimal(base_size = 14) +
         scale_fill_manual(values = clade_colors) +
         theme(axis.text.x = element_text(angle = 90, hjust = 1),
               axis.text = element_text(size = 8)) +
         # scale_y_continuous(breaks=scales::pretty_breaks(n=6)) +
         ylab("Proportion Strep group") +
         ggtitle("")

```


```{r nt_strep_proportion}

strep_prop_nt <- malt_nt %>%
  # make a column with Strep or other
  mutate(Genus = ifelse(str_detect(Species,"Strep"),"Streptococcus","Other")) %>%
  # remove Species column leaving Genus column
  select(Genus, everything()) %>%
  select(-Species) %>%
  # convert to long format
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  group_by(Library_ID, Genus) %>%
  # sum all Strep read counts and all Other read counts
  summarize(Reads = sum(Counts)) %>%
  spread(Genus, Reads) %>%
  # convert to proportions
  adorn_percentages() %>%
  ungroup()

strep_prop_nt %>%
  gather("Genus","Counts", 2:ncol(.)) %>%
  mutate(Library_ID = fct_relevel(Library_ID, sample_order_nt)) %>%
  arrange(Library_ID) %>%
  ggplot(., aes(x = Library_ID, y = Counts, fill = Genus)) +
         geom_bar(stat = "identity") + # , color = "black"
         theme_minimal(base_size = 14) +
         scale_fill_manual(values = genus_colors) +
         theme(axis.text.x = element_text(angle = 90, hjust = 1),
               axis.text = element_text(size = 8)) +
         # scale_y_continuous(breaks=scales::pretty_breaks(n=6)) +
         ylab("Proportion Strep group") +
         ggtitle("")

```

## MetaPhlAn3
Note that MetaPhlAn3 didn't pick up any strep outside of sanguinis and mitis
groups. This means that for several samples, no Strep were detected.
```{r load_mpa3_data}
# read in the species table
mpa3_raw <- fread("./05-results.backup/mpa3_abundance_table.tsv") %>%
  select(-NCBI_tax_id)

# clean the file names
colnames(mpa3_raw) <- gsub(".metaphlan3","", colnames(mpa3_raw))

# filter to have only species
mpa3_sp <- mpa3_raw %>% 
  filter(str_detect(clade_name, "s__")) %>%
  separate(clade_name, into = c("K","P","C","O","F","G","Species"), sep = "\\|") %>%
  select(-c("K","P","C","O","F","G")) %>%
  mutate(Species = str_replace_all(Species, "s__",""),
         Species = str_replace_all(Species, "_"," ")) %>%
  # these samples have no metadata
  select(-c("MID025.A0101","MID033.A0101","MID052.A0101","MID056.A0101","MID065.A0101","MID068.A0101","MID076.A0101","MID078.A0101")) %>%
  arrange(Species)

```


Now decontam the tables
```{r mpa3_decontam}
mpa3_contaminants <- fread("./05-results.backup/contaminant_species_mid_mpa3.tsv", sep = "\t") %>%
  bind_rows(., fread("./05-results.backup/contaminant_species_kilteasheen_mpa3.tsv", sep = "\t")) %>%
  bind_rows(., fread("./05-results.backup/contaminant_species_iberia_mpa3.tsv", sep = "\t")) %>%
  unique()

# remove contaminant species from nt table
mpa3_species_decontam <- mpa3_sp %>%
  anti_join(., mpa3_contaminants)

mpa3 <- mpa3_species_decontam %>%
  select(matches("Species|MID|CMB")) %>%
  adorn_totals(where = "col")  %>%
  filter(Total >  0) %>%
  select(-Total)


```

```{r mpa3_graphs}
sample_order_mpa3 <- mpa3 %>%
  filter(str_detect(Species, "Streptococcus")) %>%
  # join with the Strep grup table for plotting
  inner_join(., strep_groups) %>%
  select(-Species) %>%
  select(dRep, everything()) %>%
  # transpose
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  group_by(Library_ID, dRep) %>%
  summarize(Reads = sum(Counts)) %>%
  spread(dRep, Reads) %>%
  # convert table to proportions
  adorn_percentages(denominator = "row") %>%
  # arrange by Sanguinis and for the ones w/o Sanguinis, by anginosis
  arrange(desc(Sanguinis)) %>%
  pull(Library_ID)


mpa3 %>%
  filter(str_detect(Species, "Streptococcus")) %>%
  # join with the Strep grup table for plotting
  inner_join(., strep_groups) %>%
  select(Species, dRep, everything()) %>%
  # convert to long-format
  gather("Library_ID","Counts", 3:ncol(.)) %>%
  mutate(Library_ID = fct_relevel(Library_ID, sample_order_mpa3))%>%
  dplyr::group_by(Library_ID, dRep) %>%
  summarize(Count_sum = sum(Counts)) %>%
  ggplot(., aes(x = Library_ID, y = Count_sum, fill = dRep)) +
         geom_bar(position = "fill", stat = "identity") + # , color = "black"
         theme_minimal(base_size = 14) +
         scale_fill_manual(values = clade_colors) +
         theme(axis.text.x = element_text(angle = 90, hjust = 1),
               axis.text = element_text(size = 8)) +
         # scale_y_continuous(breaks=scales::pretty_breaks(n=6)) +
         ylab("Proportion Strep group") +
         ggtitle("")

```



```{r mpa3_strep_proportion}

strep_prop_mpa3 <- mpa3_raw %>%
  filter(str_detect(clade_name, "g__"),
         !str_detect(clade_name, "s__")) %>%
  filter(str_detect(clade_name, "Streptococcus")) %>%
  pivot_longer(!clade_name, names_to = "Library_ID", values_to = "Streptococcus") %>%
  mutate(Other = 100 - Streptococcus) %>%
  select(-clade_name) %>%
  filter(str_detect(Library_ID, "MID|CMB")) %>%
  pivot_longer(!Library_ID, names_to = "Genus", values_to = "Percent")


strep_prop_mpa3 %>%
  filter(!str_detect(Library_ID, "MID025.A0101|MID033.A0101|MID052.A0101|MID056.A0101|MID065.A0101|MID068.A0101|MID076.A0101|MID078.A0101")) %>%
  mutate(Library_ID = fct_relevel(Library_ID, sample_order_mpa3)) %>%
  arrange(Library_ID) %>%
  ggplot(., aes(x = Library_ID, y = Percent, fill = Genus)) +
         geom_bar(stat = "identity") + # , color = "black"
         theme_minimal(base_size = 14) +
         scale_fill_manual(values = genus_colors) +
         theme(axis.text.x = element_text(angle = 90, hjust = 1),
               axis.text = element_text(size = 8)) +
         # scale_y_continuous(breaks=scales::pretty_breaks(n=6)) +
         ylab("Proportion Strep group") +
         ggtitle("")

```

```{r save_mpa3_order, eval = F}

sample_order_mpa3 %>%
  as.data.frame(.) %>%
 rename(Library_ID = ".") %>%
fwrite(., "./05-results.backup/mpa_Strep_group_order.tsv", sep = "\t", quote = F)

```























