---
title: "MID/CMB, Radcliffe, Kilteasheen, Iberian medieval calculus MALT decontam"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---

```{r load_libraries, echo = F, message = F}
library(knitr)
library(decontam)
library(data.table)
library(janitor)
library(tidyverse)
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r load_data}
# read in the species table
sub10M_raw <- fread("./05-results.backup/malt_sub10M-species.txt")
sub75bp_raw <- fread("./05-results.backup/malt_sub75bp-species.txt")

# clean the file names
sub10M_raw <- rename(sub10M_raw, Species = `#Datasets`)
colnames(sub10M_raw) <- gsub(".unmapped","", colnames(sub10M_raw))
colnames(sub10M_raw) <- gsub(".10M","", colnames(sub10M_raw))

# remove human and unassigned reads, convert NA to 0, and remove the 8 MID samples with no metadata
sub10M_raw <- sub10M_raw %>% 
  filter(!str_detect(Species, "Homo sapiens|Not assigned")) %>%
  replace(is.na(.), 0) %>%
  # these samples have no metadata
  select(-c("MID025.A0101","MID033.A0101","MID052.A0101","MID056.A0101","MID065.A0101","MID068.A0101","MID076.A0101","MID078.A0101")) %>%
  arrange(Species)

sub75bp_raw <- rename(sub75bp_raw, Species = `#Datasets`)
colnames(sub75bp_raw) <- gsub(".unmapped.75bp","", colnames(sub75bp_raw))

# remove human and unassigned reads, convert NA to 0, and remove the 8 MID samples with no metadata
sub75bp_raw <- sub75bp_raw %>% 
  filter(!str_detect(Species, "Homo sapiens|Not assigned")) %>%
  replace(is.na(.), 0) %>%
  # these samples have no metadata
  select(-c("MID025.A0101","MID033.A0101","MID052.A0101","MID056.A0101","MID065.A0101","MID068.A0101","MID076.A0101","MID078.A0101")) %>%
  arrange(Species)

```

Load in the original table to pull the LIB and EXB samples
```{r}
load("./05-results.backup/Taxonomy_tables.RData")

```

```{r}
# merge the EXB and LIB samples from here with the raw tables

sub10M_raw <- sub10M_raw %>%
  full_join(., all_species_raw %>%
              select(matches("Species|EXB|LIB")), by = "Species") %>%
  replace(is.na(.), 0) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)
   
sub75bp_raw <- sub75bp_raw %>%
  full_join(., all_species_raw %>%
              select(matches("Species|EXB|LIB")), by = "Species") %>%
  replace(is.na(.), 0) %>%
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total)
 
```


Load metadata
```{r load_metadata}

load("./05-results.backup/Metadata_tables.RData")

```


Read in the DeepEvo table with ARS samples, to use as bone controls to hopefully
clean up more environmental taxa. From
`https://github.com/jfy133/Hominid_Calculus_Microbiome_Evolution/tree/master/06-additional_data_files/Data_R11_S12_MALTMEGANTables/`.
```{r}
ars_bone_sp <- fread("./05-results.backup/Evolution-Comparison_MEGAN_20190410-ex_absolute_species_prokaryotes_summarised_refseq.txt") %>%
  rename(Species = `#Datasets`) %>%
  column_to_rownames("Species") %>%
  select(matches("ARS")) %>%
  rownames_to_column("Species") %>%
# remove all species that aren't in the gut samples
  adorn_totals(where = "col") %>%
  filter(Total != 0) %>%
  select(-Total)

colnames(ars_bone_sp) <- gsub("_S0_L003_R1_001.fastq.combined.fq.prefixed.extractunmapped.bam","", colnames(ars_bone_sp))

```

```{r deep_evo_meta}
deep_evo_metadata <- fread("./00-documentation.backup/02-calculus_microbiome-deep_evolution-individualscontrolssources_metadata_20200219.tsv") %>%
  rename(Library_ID = `#SampleID`) %>%
  filter(str_detect(Library_ID, "ARS"))

ars_samples <- deep_evo_metadata %>%
  pull(Library_ID)
  
```


```{r pca_plotting_functions}
# plotting PCA with colored dots
plot_pca <- function(df, pc1, pc2, metadata_group) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))
    metadata_group_shapes <- get(paste(metadata_group, "_shapes", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("Library_ID") %>%
              inner_join(full_metadata %>%
                           select(Library_ID, Site, Sample_or_Control), by = "Library_ID")
    
    metadata_group = df_X[[metadata_group]]
    
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                           xaxis <- c("PC1")
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                           xaxis <- c("PC2")
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                           xaxis <- c("PC3")
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                           yaxis <- c("PC1")
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                           yaxis <- c("PC2")
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                           yaxis <- c("PC3")
                       }

    pca_plot <- ggplot(df_X, aes(pc1, pc2, colour = metadata_group, shape = metadata_group)) +
                        geom_point(size = 4) +
                        scale_colour_manual(values = metadata_group_colors) +
                        scale_shape_manual(values = metadata_group_shapes) +
                        # stat_ellipse() +
                        xlab(paste(xaxis, " - ", exp_var_pc1)) +
                        ylab(paste(yaxis, " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot)
}

# this function plots the PCA with the sample names instead of points
plot_pca_names <- function(df, pc1, pc2, metadata_group) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("Library_ID") %>%
              inner_join(full_metadata %>%
                           select(Library_ID, Site, Sample_or_Control), by = "Library_ID")
    
    metadata_group = df_X[[metadata_group]]
    
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                           xaxis <- c("PC1")
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                           xaxis <- c("PC2")
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                           xaxis <- c("PC3")
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                           yaxis <- c("PC1")
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                           yaxis <- c("PC2")
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                           yaxis <- c("PC3")
                       }

    pca_plot_names <- ggplot(df_X, aes(pc1, pc2, colour = metadata_group)) +
                        geom_text(aes(label = Library_ID), size = 2) +
                        scale_colour_manual(values = metadata_group_colors) +
                        # stat_ellipse() +
                        xlab(paste(xaxis, " - ", exp_var_pc1)) +
                        ylab(paste(yaxis, " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot_names)
}

```


### sub10M
Before running decontam, let's plot the samples in a PCA to see if there are any that are clearly not good-quality
```{r raw_plot_sub10M}

malt_species_raw <- sub10M_raw %>%
  gather("Library_ID","Counts",2:ncol(.)) %>%
  mutate(Counts = Counts + 1) %>%
  spread(Species,Counts) %>%
  column_to_rownames("Library_ID")

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.malt_species_pca <- mixOmics::tune.pca(malt_species_raw, logratio = 'CLR')
tune.malt_species_pca

# perform a PCA to see how the data cluster
malt_species.pca <- mixOmics::pca(malt_species_raw, ncomp = 3, logratio = 'CLR')
plot(malt_species.pca)

# define some colors
Sample_or_Control_colors = c("Blue", "Magenta")
Sample_or_Control_shapes = c(19,8)

plot_pca(malt_species.pca, "PC1", "PC2", "Sample_or_Control")
plot_pca_names(malt_species.pca,  "PC1", "PC2", "Sample_or_Control")

```

```{r outliers_10M}
# these are the 3 blanks that "passed" the cuperdec cut-off
# and we can see they plot with samples in the PCA above
outliers_sub10M <- c("EXB059.A2501","LIB058.A0103","LIB058.A0106","CSS","CSD")
poor_samples_10M <- fread("./05-results.backup/cuperdec_sub10M_poor_samples.tsv") %>%
  pull(Sample)

```

### sub75bp
Before running decontam, let's plot the samples in a PCA to see if there are any that are clearly not good-quality
```{r raw_plot_sub10M}

malt_species_raw <- sub75bp_raw %>%
  gather("Library_ID","Counts",2:ncol(.)) %>%
  mutate(Counts = Counts + 1) %>%
  spread(Species,Counts) %>%
  column_to_rownames("Library_ID")

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.malt_species_pca <- mixOmics::tune.pca(malt_species_raw, logratio = 'CLR')
tune.malt_species_pca

# perform a PCA to see how the data cluster
malt_species.pca <- mixOmics::pca(malt_species_raw, ncomp = 3, logratio = 'CLR')
plot(malt_species.pca)

# define some colors
Sample_or_Control_colors = c("Blue", "Magenta")
Sample_or_Control_shapes = c(19,8)

plot_pca(malt_species.pca, "PC1", "PC2", "Sample_or_Control")
plot_pca_names(malt_species.pca,  "PC1", "PC2", "Sample_or_Control")

```

```{r outlier_75bps}
# these are the 3 blanks that "passed" the cuperdec cut-off
# and we can see they plot with samples in the PCA above
outliers_sub75bp <- c("EXB059.A2501","LIB058.A0103","LIB058.A0106","CSS","CSD")
poor_samples_75bp <- fread("./05-results.backup/cuperdec_sub75bp_poor_samples.tsv") %>%
  pull(Sample)

```

# Decontamination with Decontam
We want to know if there are any species that can be removed b/c they appear to
be contaminants. Use the package decontam to determine which species may be
contaminants. Prepare the matrix for decontam.


## sub10M
```{r decontam_data_prep}
# remove NA value samples from the input data frame
malt_species_10M.dataframe <- sub10M_raw %>%
  filter(!str_detect(Species, "Not assigned|Homo sapiens")) %>%
  select(-one_of(outliers_sub10M)) %>%
  select(-one_of(poor_samples_10M)) %>%
  full_join(., ars_bone_sp) %>%
  replace(is.na(.), 0) %>%
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  spread(Species,Counts) 

# make a matrix for decontam
malt_species.matrix <- data.matrix(malt_species.dataframe[2:ncol(malt_species.dataframe)])
rownames(malt_species.matrix) <- malt_species.dataframe$Library_ID

# make a metadata dataframe with the same NA-containing samples removed
metadata.decontam <- full_metadata %>%
  filter(!str_detect(Library_ID, poor_samples_10M %>% str_c(collapse = "|"))) %>%
  filter(!str_detect(Library_ID, outliers_sub10M %>% str_c(collapse = "|"))) %>%
  filter(!str_detect(Library_ID, "VLC")) %>% 
  # these samples have no metadata
  filter(!str_detect(Library_ID, "LIB058.A0101|MID025.A0101|MID033.A0101|MID052.A0101|MID056.A0101|MID065.A0101|MID068.A0101|MID076.A0101|MID078.A0101|CSN|CSL")) %>%
  arrange(Library_ID) %>%
  distinct() %>%
  bind_rows(., deep_evo_metadata %>%
              mutate(Site = "Bone") %>%
              select(Site, Library_ID, Sample_or_Control)) %>%
  arrange(Site)

metadata.decontam %>% select(Library_ID) %>%
  anti_join(., malt_species.dataframe %>% select(Library_ID))

```

Use the prevalence method for all but Kilteasheen

### MID/CMB

```{r decontam_data_prep}
# remove NA value samples from the input data frame
malt_species_10M_mid.dataframe <- sub10M_raw %>%
  filter(!str_detect(Species, "Not assigned|Homo sapiens")) %>%
  select(-one_of(outliers_sub10M)) %>%
  select(-one_of(poor_samples_10M)) %>%
  full_join(., ars_bone_sp) %>%
  select(matches("Species|ARS|MID|CMB|LIB058.A01|EXB059|EXB034")) %>%
  replace(is.na(.), 0) %>%
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  spread(Species,Counts) %>%
  arrange(Library_ID)

# make a matrix for decontam
malt_species_10M_mid.matrix <- data.matrix(malt_species_10M_mid.dataframe[2:ncol(malt_species_10M_mid.dataframe)])
rownames(malt_species_10M_mid.matrix) <- malt_species_10M_mid.dataframe$Library_ID

# make a metadata dataframe with the same NA-containing samples removed
metadata_mid.decontam <- full_metadata %>%
  filter(!str_detect(Library_ID, poor_samples_10M %>% str_c(collapse = "|"))) %>%
  filter(!str_detect(Library_ID, outliers_sub10M %>% str_c(collapse = "|"))) %>%
  filter(str_detect(Library_ID, "ARS|MID|CMB|LIB058.A01|EXB059|EXB034")) %>%
  # these samples have no metadata
  filter(!str_detect(Library_ID, "LIB058.A0101|MID025.A0101|MID033.A0101|MID052.A0101|MID056.A0101|MID065.A0101|MID068.A0101|MID076.A0101|MID078.A0101|CSN|CSL")) %>%
  arrange(Library_ID) %>%
  distinct() %>%
  bind_rows(., deep_evo_metadata %>%
              mutate(Site = "Bone") %>%
              select(Site, Library_ID, Sample_or_Control)) %>%
  arrange(Site)

metadata_mid.decontam %>% select(Library_ID) %>%
  anti_join(., malt_species_10M_mid.dataframe %>% select(Library_ID))

```

```{r decontam_prevalence}
# Assign sample-type status to identify control samples
metadata.decontam$is.control <- metadata.decontam$Sample_or_Control=="Control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.species <- isContaminant(malt_species.matrix, method="prevalence", 
                                 neg=metadata.decontam$is.control, 
                                 threshold = 0.55)

# Check number of contaminants
table(contam.prev.species$contaminant)

# List of contaminants
cont_prev_species_list <- contam.prev.species[which(contam.prev.species$contaminant=="TRUE"), ]

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')


```

```{r contaminants, eval = F}
contaminant_species_sub10M <- cont_prev_species_list %>% rownames_to_column("Species") %>% select(Species)
fwrite(contaminant_species_sub10M, file = "./05-results.backup/contaminant_species_sub10M.tsv", sep="\t", quote=F)

```

## sub75bp
```{r decontam_data_prep}
# list out the samples that have all 0 or NA values to remove from the input matrix
# NA_filter_list <- metadata %>% filter((is.na(SumLibraryConcentration_copies_ul) | SumLibraryConcentration_copies_ul <= 0)) %>% select(Library_ID)

# remove NA value samples from the input data frame
malt_species.dataframe <- sub75bp_raw %>%
  filter(!str_detect(Species, "Not assigned|Homo sapiens")) %>%
  select(-one_of(outliers_sub75bp)) %>%
  select(-one_of(poor_samples_75bp)) %>%
  full_join(., ars_bone_sp) %>%
  replace(is.na(.), 0) %>%
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  spread(Species,Counts) 
# %>%
#   anti_join(., NA_filter_list)

# make a matrix for decontam
malt_species.matrix <- data.matrix(malt_species.dataframe[2:ncol(malt_species.dataframe)])
rownames(malt_species.matrix) <- malt_species.dataframe$Library_ID

# make a metadata dataframe with the same NA-containing samples removed
metadata.decontam <- full_metadata %>%
  filter(!str_detect(Library_ID, poor_samples_75bp %>% str_c(collapse = "|"))) %>%
  # anti_join(., NA_filter_list, by = "Library_ID") %>%
  filter(!str_detect(Library_ID, outliers_sub75bp %>% str_c(collapse = "|"))) %>%
  filter(Library_ID != "LIB058.A0101") %>% # this sample is not in the metadata
  filter(!str_detect(Library_ID, "MID025.A0101|MID033.A0101|MID052.A0101|MID056.A0101|MID065.A0101|MID068.A0101|MID076.A0101|MID078.A0101|CSN|CSL")) %>%
  arrange(Library_ID) %>%
  distinct() %>%
  bind_rows(., deep_evo_metadata %>%
              mutate(Site = "Bone") %>%
              select(Site, Library_ID, Sample_or_Control)) %>%
  arrange(Site)

metadata.decontam %>% select(Library_ID) %>%
  anti_join(., malt_species.dataframe %>% select(Library_ID))

```

Try the prevalence method to compare. 
```{r decontam_prevalence, eval = F}
# Assign sample-type status to identify control samples
metadata.decontam$is.control <- metadata.decontam$Sample_or_Control=="Control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.species <- isContaminant(malt_species.matrix, method="prevalence", 
                                 neg=metadata.decontam$is.control, 
                                 threshold = 0.6)

# Check number of contaminants
table(contam.prev.species$contaminant)

# List of contaminants
cont_prev_species_list_sub75bp <- contam.prev.species[which(contam.prev.species$contaminant=="TRUE"), ]

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')


```

```{r contaminants, eval = F}
contaminant_species_sub75bp <- cont_prev_species_list_sub75bp %>% rownames_to_column("Species") %>% select(Species)
fwrite(contaminant_species_sub75bp, file = "./05-results.backup/contaminant_species_sub75bp.tsv", sep="\t", quote=F)

```















