---
title: "Cameroon hunter-gatherer calculus/plaque MALT decontam"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
    github_document:
        fig_width: 8
        fig_height: 6
        toc: true
---

```{r load_libraries, echo = F, message = F}
library(knitr)
library(decontam)
library(data.table)
library(janitor)
library(tidyverse)
opts_chunk$set(echo=F, warning=F, message=F, dpi=150)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r load_data}
mid_raw <- fread("./05-results.backup/MID_malt_bactarchhomo2018_comparison-species.txt")
radcliffe_raw <- fread("./05-results.backup/Radcliffe_pfuT_comparison-species.txt")
kilteasheen_raw <- fread("./05-results.backup/Kilteasheen_comparison-species.txt")
iberian_raw <- fread("./05-results.backup/iberian_medieval_malt_species_raw.tsv")


all_raw <- mid_raw %>%
  full_join(., radcliffe_raw) %>%
  full_join(., kilteasheen_raw) %>%
  full_join(., iberian_raw)


# clean the file names
all_raw <- rename(all_raw, Species = `#Datasets`)
colnames(all_raw) <- gsub(".unmapped","", colnames(all_raw))
colnames(all_raw) <- gsub("MeganServer::","", colnames(all_raw))

# remove human and unassigned reads, convert NA to 0, and remove the 8 MID samples with no metadata
all_raw <- all_raw %>% 
  filter(!str_detect(Species, "Homo sapiens|Not assigned")) %>%
  replace(is.na(.), 0) %>%
  # these samples have no metadata
  select(-c("MID025.A0101","MID033.A0101","MID052.A0101","MID056.A0101","MID065.A0101","MID068.A0101","MID076.A0101","MID078.A0101")) %>%
  # remove Radcliffe soil and dentin
  select(-c("CSS","CSD")) %>%
  # remove any taxa that are not present in the remaining samples
  adorn_totals(where = "col") %>%
  filter(Total > 0) %>%
  select(-Total) %>%
  arrange(Species)

```

```{r}
metadata <- fread("00-documentation.backup/MID_analysis_metadata.tsv") %>%
  select(Site, Library_ID, Sample_or_Control) %>%
  full_join(., fread("./00-documentation.backup/radcliffe_pfuT_metadata.tsv") %>%
            select(Site, Library_ID, Sample_or_Control)) %>%
  full_join(., fread("./00-documentation.backup/kilteasheen_metadata.tsv") %>%
            select(Site, Library_ID, Sample_or_Control)) %>%
  full_join(., fread("./00-documentation.backup/iberian_medieval_metadata.tsv") %>%
            select(Site, Library_ID, Sample_or_Control))

#set the columns of interest to factors
metadata <- metadata %>%
  as_tibble() %>%
  mutate(Site = fct_relevel(Site, "MID","CMB","Radcliffe","Kilteasheen","ELR",
                                "IVE","Extraction_blank","Library_blank"),
         Sample_or_Control = fct_relevel(Sample_or_Control, "Sample", "Control"))

```


```{r pca_plotting_functions}
# plotting PCA with colored dots
plot_pca <- function(df, pc1, pc2, metadata_group) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))
    metadata_group_shapes <- get(paste(metadata_group, "_shapes", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("Library_ID") %>%
              inner_join(metadata %>%
                           select(Library_ID, Site, Sample_or_Control), by = "Library_ID")
    
    metadata_group = df_X[[metadata_group]]
    
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                           xaxis <- c("PC1")
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                           xaxis <- c("PC2")
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                           xaxis <- c("PC3")
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                           yaxis <- c("PC1")
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                           yaxis <- c("PC2")
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                           yaxis <- c("PC3")
                       }

    pca_plot <- ggplot(df_X, aes(pc1, pc2, colour = metadata_group, shape = metadata_group)) +
                        geom_point(size = 4) +
                        scale_colour_manual(values = metadata_group_colors) +
                        scale_shape_manual(values = metadata_group_shapes) +
                        # stat_ellipse() +
                        xlab(paste(xaxis, " - ", exp_var_pc1)) +
                        ylab(paste(yaxis, " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot)
}

# this function plots the PCA with the sample names instead of points
plot_pca_names <- function(df, pc1, pc2, metadata_group) {
    metadata_group_colors <- get(paste(metadata_group, "_colors", sep = ""))

    exp_var <- paste0(round(df$explained_variance * 100, 2), "%")
    df_X <- df$variates$X %>%
              as.data.frame() %>%
              rownames_to_column("Library_ID") %>%
              inner_join(metadata %>%
                           select(Library_ID, Site, Sample_or_Control), by = "Library_ID")
    
    metadata_group = df_X[[metadata_group]]
    
                       if (pc1 == 'PC1') {
                           pc1 <- df_X$PC1
                           exp_var_pc1 <- exp_var[1]
                           xaxis <- c("PC1")
                       }  else if (pc1 == 'PC2') {
                           pc1 <- df_X$PC2
                           exp_var_pc1 <- exp_var[2]
                           xaxis <- c("PC2")
                       } else if (pc1 == 'PC3') {
                           pc1 <- df_X$PC3
                           exp_var_pc1 <- exp_var[3]
                           xaxis <- c("PC3")
                       }

                       if (pc2 == 'PC1') {
                           pc2 <- df_X$PC1
                           exp_var_pc2 <- exp_var[1]
                           yaxis <- c("PC1")
                       }  else if (pc2 == 'PC2') {
                           pc2 <- df_X$PC2
                           exp_var_pc2 <- exp_var[2]
                           yaxis <- c("PC2")
                       } else if (pc2 == 'PC3') {
                           pc2 <- df_X$PC3
                           exp_var_pc2 <- exp_var[3]
                           yaxis <- c("PC3")
                       }

    pca_plot_names <- ggplot(df_X, aes(pc1, pc2, colour = metadata_group)) +
                        geom_text(aes(label = Library_ID), size = 2) +
                        scale_colour_manual(values = metadata_group_colors) +
                        # stat_ellipse() +
                        xlab(paste(xaxis, " - ", exp_var_pc1)) +
                        ylab(paste(yaxis, " - ", exp_var_pc2)) +
                        theme_minimal(base_size = 10) +
                        theme(legend.position = "top") +
                        theme(legend.title = element_blank())

    return(pca_plot_names)
}

```


Before running decontam, let's plot the samples in a PCA to see if there are any that are clearly not good-quality
```{r raw_plot}

malt_species_raw <- all_raw %>%
  gather("Library_ID","Counts",2:ncol(.)) %>%
  mutate(Counts = Counts + 1) %>%
  spread(Species,Counts) %>%
  column_to_rownames("Library_ID")

# prepare to run a PCA
# check the number of components to retain by tuning the PCA
tune.malt_species_pca <- mixOmics::tune.pca(malt_species_raw, logratio = 'CLR')
tune.malt_species_pca

# perform a PCA to see how the data cluster
malt_species.pca <- mixOmics::pca(malt_species_raw, ncomp = 3, logratio = 'CLR')
plot(malt_species.pca)

# define some colors
Sample_or_Control_colors = c("Blue", "Magenta")
Sample_or_Control_shapes = c(19,8)

plot_pca(malt_species.pca, "PC1", "PC2", "Sample_or_Control")
plot_pca_names(malt_species.pca,  "PC1", "PC2", "Sample_or_Control")

```

```{r outliers}
# these are the 3 blanks that "passed" the cuperdec cut-off
# and we can see they plot with samples in the PCA above
outliers <- c("EXB059.A2501","LIB058.A0103","LIB058.A0106","CSS","CSD")
poor_samples <- fread("./05-results.backup/cuperdec_poor_samples.tsv") %>%
  pull(Sample)

```

# Decontamination with Decontam
We want to know if there are any species that can be removed b/c they appear to
be contaminants. Use the package decontam to determine which species may be
contaminants. Prepare the matrix for decontam.


## All together
```{r decontam_data_prep}
# list out the samples that have all 0 or NA values to remove from the input matrix
# NA_filter_list <- metadata %>% filter((is.na(SumLibraryConcentration_copies_ul) | SumLibraryConcentration_copies_ul <= 0)) %>% select(Library_ID)

# remove NA value samples from the input data frame
malt_species.dataframe <- all_raw %>%
  filter(!str_detect(Species, "Not assigned|Homo sapiens")) %>%
  select(-one_of(outliers)) %>%
  select(-one_of(poor_samples)) %>%
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  spread(Species,Counts) 
# %>%
#   anti_join(., NA_filter_list)

# make a matrix for decontam
malt_species.matrix <- data.matrix(malt_species.dataframe[2:ncol(malt_species.dataframe)])
rownames(malt_species.matrix) <- malt_species.dataframe$Library_ID

# make a metadata dataframe with the same NA-containing samples removed
metadata.decontam <- metadata %>%
  filter(!str_detect(Library_ID, poor_samples %>% str_c(collapse = "|"))) %>%
  # anti_join(., NA_filter_list, by = "Library_ID") %>%
  filter(!str_detect(Library_ID, outliers %>% str_c(collapse = "|"))) %>%
  filter(Library_ID != "LIB058.A0101") %>% # this sample is not in the metadata
  arrange(Library_ID) %>%
  distinct()

# metadata.decontam %>% select(Library_ID) %>%
#   anti_join(., malt_species.dataframe %>% select(Library_ID))

```

Try the prevalence method to compare. 
```{r decontam_prevalence}
# Assign sample-type status to identify control samples
metadata.decontam$is.control <- metadata.decontam$Sample_or_Control=="Control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam.prev.species <- isContaminant(malt_species.matrix, method="prevalence", 
                                 neg=metadata.decontam$is.control, 
                                 threshold = 0.8)

# Check number of contaminants
table(contam.prev.species$contaminant)

# List of contaminants
cont_prev_species_list <- contam.prev.species[which(contam.prev.species$contaminant=="TRUE"), ]

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam.prev.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')


```

## MID, CMB
```{r decontam_data_prep}
# list out the samples that have all 0 or NA values to remove from the input matrix
# NA_filter_list <- metadata %>% filter((is.na(SumLibraryConcentration_copies_ul) | SumLibraryConcentration_copies_ul <= 0)) %>% select(Library_ID)

# remove NA value samples from the input data frame
malt_species_mid.dataframe <- mid_raw %>%
  rename(Species = `#Datasets`) %>%
  filter(!str_detect(Species, "Not assigned|Homo sapiens")) %>%
  select(-one_of(outliers)) %>%
  select(-one_of(poor_samples)) %>%
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  spread(Species,Counts) 
# %>%
#   anti_join(., NA_filter_list)

# make a matrix for decontam
malt_species_mid.matrix <- data.matrix(malt_species_mid.dataframe[2:ncol(malt_species_mid.dataframe)])
rownames(malt_species_mid.matrix) <- malt_species_mid.dataframe$Library_ID

# make a metadata dataframe with the same NA-containing samples removed
metadata_mid.decontam <- metadata %>%
  filter(!str_detect(Library_ID, poor_samples %>% str_c(collapse = "|"))) %>%
  # anti_join(., NA_filter_list, by = "Library_ID") %>%
  filter(!str_detect(Library_ID, outliers %>% str_c(collapse = "|"))) %>%
  filter(str_detect(Library_ID, "MID|CMB|EXB059|LIB058")) %>%
  filter(Library_ID != "LIB058.A0101") %>% # this sample is not in the metadata
  arrange(Library_ID) %>%
  distinct()

# metadata.decontam %>% select(Library_ID) %>%
#   anti_join(., malt_species.dataframe %>% select(Library_ID))

```

```{r decontam_prevalence}
# Assign sample-type status to identify control samples
metadata_mid.decontam$is.control <- metadata_mid.decontam$Sample_or_Control=="Control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam_mid.prev.species <- isContaminant(malt_species_mid.matrix, method="prevalence", 
                                 neg=metadata_mid.decontam$is.control, 
                                 threshold = 0.8)

# Check number of contaminants
table(contam_mid.prev.species$contaminant)

# List of contaminants
cont_prev_species_list_mid <- contam_mid.prev.species[which(contam_mid.prev.species$contaminant=="TRUE"), ]

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam_mid.prev.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')


```

## Radcliffe
```{r decontam_data_prep}
# list out the samples that have all 0 or NA values to remove from the input matrix
# NA_filter_list <- metadata %>% filter((is.na(SumLibraryConcentration_copies_ul) | SumLibraryConcentration_copies_ul <= 0)) %>% select(Library_ID)

# remove NA value samples from the input data frame
malt_species_rad.dataframe <- radcliffe_raw %>%
  rename(Species = `#Datasets`) %>%
  filter(!str_detect(Species, "Not assigned|Homo sapiens")) %>%
  select(-one_of(outliers)) %>%
  select(-one_of(poor_samples)) %>%
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  spread(Species,Counts) 
# %>%
#   anti_join(., NA_filter_list)

# make a matrix for decontam
malt_species_rad.matrix <- data.matrix(malt_species_rad.dataframe[2:ncol(malt_species_rad.dataframe)])
rownames(malt_species_rad.matrix) <- malt_species_rad.dataframe$Library_ID

# make a metadata dataframe with the same NA-containing samples removed
metadata_rad.decontam <- metadata %>%
  filter(!str_detect(Library_ID, poor_samples %>% str_c(collapse = "|"))) %>%
  # anti_join(., NA_filter_list, by = "Library_ID") %>%
  filter(!str_detect(Library_ID, outliers %>% str_c(collapse = "|"))) %>%
  filter(str_detect(Library_ID, "CS")) %>%
  arrange(Library_ID) %>%
  distinct()

# metadata.decontam %>% select(Library_ID) %>%
#   anti_join(., malt_species.dataframe %>% select(Library_ID))

```

```{r decontam_prevalence}
# Assign sample-type status to identify control samples
metadata_rad.decontam$is.control <- metadata_rad.decontam$Sample_or_Control=="Control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam_rad.prev.species <- isContaminant(malt_species_rad.matrix, method="prevalence", 
                                 neg=metadata_rad.decontam$is.control, 
                                 threshold = 0.8)

# Check number of contaminants
table(contam_rad.prev.species$contaminant)

# List of contaminants
cont_prev_species_list_rad <- contam_rad.prev.species[which(contam_rad.prev.species$contaminant=="TRUE"), ]

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam_rad.prev.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')


```

## Iberian medieval
```{r decontam_data_prep}
# list out the samples that have all 0 or NA values to remove from the input matrix
# NA_filter_list <- metadata %>% filter((is.na(SumLibraryConcentration_copies_ul) | SumLibraryConcentration_copies_ul <= 0)) %>% select(Library_ID)

# remove NA value samples from the input data frame
malt_species_iber.dataframe <- iberian_raw %>%
  rename(Species = `#Datasets`) %>%
  filter(!str_detect(Species, "Not assigned|Homo sapiens")) %>%
  select(-one_of(outliers)) %>%
  select(-one_of(poor_samples)) %>%
  gather("Library_ID","Counts", 2:ncol(.)) %>%
  spread(Species,Counts) 
# %>%
#   anti_join(., NA_filter_list)

# make a matrix for decontam
malt_species_iber.matrix <- data.matrix(malt_species_iber.dataframe[2:ncol(malt_species_iber.dataframe)])
rownames(malt_species_iber.matrix) <- malt_species_iber.dataframe$Library_ID

# make a metadata dataframe with the same NA-containing samples removed
metadata_iber.decontam <- metadata %>%
  filter(!str_detect(Library_ID, poor_samples %>% str_c(collapse = "|"))) %>%
  # anti_join(., NA_filter_list, by = "Library_ID") %>%
  filter(!str_detect(Library_ID, outliers %>% str_c(collapse = "|"))) %>%
  filter(str_detect(Library_ID, "ELR|IVE|EXB015|LIB007")) %>%
  arrange(Library_ID) %>%
  distinct()

# metadata.decontam %>% select(Library_ID) %>%
#   anti_join(., malt_species.dataframe %>% select(Library_ID))

```

```{r decontam_prevalence}
# Assign sample-type status to identify control samples
metadata_iber.decontam$is.control <- metadata_iber.decontam$Sample_or_Control=="Control"

#Try changing the probability threshold cut-off to see how many more or less are ID'd as contaminants
contam_iber.prev.species <- isContaminant(malt_species_iber.matrix, method="prevalence", 
                                 neg=metadata_iber.decontam$is.control, 
                                 threshold = 0.8)

# Check number of contaminants
table(contam_iber.prev.species$contaminant)

# List of contaminants
cont_prev_species_list_iber <- contam_iber.prev.species[which(contam_iber.prev.species$contaminant=="TRUE"), ]

# Look at the histogram - does the probability threshold need to be changed?
ggplot(contam_iber.prev.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Prevalence - Species')


```

## Kilteasheen
?? No controls available

```{r}
# Overlapping contaminant species
cont_prev_species_list_mid %>% rownames_to_column("Species") %>%
  select(Species) %>% mutate(MID = "Yes") %>%
  full_join(., cont_prev_species_list_rad %>% rownames_to_column("Species") %>%
            select(Species) %>% mutate(Radcliffe = "Yes")) %>%
  full_join(., cont_prev_species_list_iber %>% rownames_to_column("Species") %>%
            select(Species) %>% mutate(Iberia = "Yes")) %>%
  mutate(Shared = ifelse(MID == "Yes" & Radcliffe == "Yes" & Iberia == "Yes", "M-R-I", 
                         ifelse(MID == Radcliffe & is.na(Iberia), "M-R",
                                ifelse(MID == "Yes" & Iberia == "Yes" & is.na(Radcliffe), "M-I",
                                       ifelse(Radcliffe == "Yes" & Iberia == "Yes" & is.na(MID), "R-I", "Single")))))

```


We'll use the prevalence method, because this doesn't include known oral taxa in the contaminant list (frequency method does), and it picks up a lot more than the combined method does (but is that important if we cut off the tail of low-abundance taxa (say <0.01% abundance) anyway?). But which cut-off to use? How does changing the cut-off affect the tail of low-abundance taxa?
List out the pathways that are considered contaminants, to be removed from the table prior to all analyses. Then remove them from the table to make the decontaminated input table for analysis. *Note that the genus list includes Actinomyces, and this needs to be removed from the list b/c Actinomyces are a well known oral genus*
```{r contaminants, eval = F}
contaminants_species <- cont_prev_species_list %>% rownames_to_column("Species") %>% select(Species)
malt_species.decontam <- malt_species %>%
  anti_join(., contaminants_species)

write.table(malt_species.decontam, file = "05-results.backup/malt_species.decontam.tsv", sep="\t", quote=F, row.names = F)
write.table(malt_species, "05-results.backup/malt_species.tsv", sep="\t", quote=F, row.names = F)
write.table(contaminants_species, "05-results.backup/malt_contaminants_species.tsv", sep="\t", quote=F, row.names = F)

contaminants_genus <- cont_prev_genus_list %>% rownames_to_column("Species") %>% select(Species) %>% filter(Species != "Actinomyces")
malt_genus.decontam <- malt_genus %>%
  anti_join(., contaminants_genus)

write.table(contaminants_genus, "05-results.backup/malt_contaminants_genus.tsv", sep="\t", quote=F, row.names = F)

```

## Not used

Now we're ready to run Decontam. Start with the frequency method. 
```{r decontam_frequency, eval = F}
# Check for contamination by fequency
contam.freq.species <- isContaminant(malt_species.matrix, method="frequency", 
                                 conc=metadata.decontam$SumLibraryConcentration_copies_ul, 
                                 threshold = 0.9)

# Print the # of species identified as contaminants by frequency
table(contam.freq.species$contaminant)

cont_freq_species_list <- contam.freq.species[which(contam.freq.species$contaminant=="TRUE"), ]

# Make a histogram to look at the distribution of the p-value (probability for classifying contaminants)

ggplot(contam.freq.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  ggtitle('Frequency - Species')

```

Check the combined prevalence and frequency method to see how it differs from the individual methods. 
```{r decontam_combined, eval=F}
# Assign sample-type status to identify control samples (same as was done for prevalence)
metadata.decontam$is.control <- metadata.decontam$Sample_or_Control=="Control"

# Assign contaminant status
contam.comb.species <- isContaminant(malt_species.matrix, method="combined", 
                                 conc=metadata.decontam$SumLibraryConcentration_copies_ul, 
                                 neg=metadata.decontam$is.control,
                                 threshold = 0.9)

## Check how many contaminants
table(contam.comb.species$contaminant)

## List of contaminants
cont_comb_species_list <- contam.comb.species[which(contam.comb.species$contaminant=="TRUE"), ]

# Graph it to look at the histogram and pick a suitable probability-values cut-off
ggplot(contam.comb.species, aes(x=p)) +
  geom_histogram(bins=100) +
  scale_y_log10() +
  geom_hline(yintercept=10, linetype="dashed", color = "red") +
  geom_hline(yintercept=2, linetype="dashed", color = "red") +
  ggtitle('Combined - Species')

```
















