---
title: "Smokers calculus horseshoe investigation"
author: "Irina Velsko"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output: html_notebook
---

```{r load_libraries, echo = F, message = F}
library(knitr)
library(mixOmics)
library(data.table)
library(vegan)
library(compositions)
library(janitor)
library(tidyverse)
library(cowplot)
```

Set the working directory of the notebook to the parent directory of the 
notebook itself.  

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r load_data}
# malt_otu <- fread("~/Desktop/malt_otu_cuperdec_and_library_for_beta_cleaned.tsv")

# load the taxonomy tables that have been decontaminated and the column headers are cleaned up
load("./05-results.backup/Taxonomy_tables.RData")


```

Read in the metadata tables
```{r metadata}
# load the metadata file
load("./05-results.backup/Metadata_tables.RData")

```

List the outliers from the plot in `MID_CMB_Rad_Kil_Iber_decontam.Rmd`. 
```{r outliers}
outliers <- c("EXB059.A2501","LIB058.A0103","LIB058.A0106","CSS","CSD")
poor_samples <- fread("./05-results.backup/cuperdec_poor_samples.tsv") %>%
  pull(Sample)

outliersF <- str_c(outliers, collapse = "|")

```

## Unfiltered data
```{r}
malt_otu_pca_table <- all_species_decontam %>%
  select(matches("Species|MID")) %>%
  select(-one_of(outliers)) %>%
  select(-one_of(poor_samples)) %>%
  gather("Library_ID","Counts",2:ncol(.)) %>%
  mutate(Counts = Counts + 1) %>%
  spread(Species, Counts) %>%
  column_to_rownames("Library_ID")

# look at the screenplot to decide how many components to keep for the PCA -> will keep all but 2
mixOmics::tune.pca(malt_otu_pca_table, logratio = 'CLR')
# perform a PCA to see how the data cluster
malt_otu.pca <- mixOmics::pca(malt_otu_pca_table, ncomp = 2, logratio = 'CLR')
malt_pca_values <- malt_otu.pca$variates$X %>%
  as.data.frame() %>%
  rownames_to_column("Library_ID") %>%
  inner_join(., full_metadata, by = "Library_ID")

```
There is a point, point where bars jump from tall to much shorter, this number is used to determine numbers of components for pca (2 bars high, dropdown = leave at 2, only one bar = leave at 2), these contain most contaminations. Will leave 2 as number.

Plot the PCA
```{r}
# plot the PCA with Library_IDs
ggplot(malt_pca_values, aes(PC1, PC2, color = Region, shape = Region)) +
  geom_text(aes(label = Library_ID), size = 3) +
  theme_minimal()

horseshoe <- ggplot(malt_pca_values, aes(PC1, PC2, color = Region, shape = Region)) +
  geom_point(size = 2) +
  theme_minimal()

horseshoe

# ggsave("~/Desktop/MID_horseshoe.pdf", plot = horseshoe, device = "pdf",
#        scale = 1, width = 7, height = 5, units = c("in"), dpi = 300)

```



```{r}

# plot with the species ordered by loading in PC1
# Plot the heatmap with samples ordered by the loading in PC1 from above
# make the Library_IDs a factor so they're in the correct order for the heatmap using the vector from the step above


sample_order <- malt_pca_values %>%
  arrange(PC1) %>%
  pull(Library_ID)

species_order <- malt_otu.pca$loadings$X %>%
  as.data.frame() %>%
  rownames_to_column("Species") %>%
  arrange(desc(PC1)) %>%
  filter(!between(PC1, -0.01, 0.01)) %>%
  pull(Species)

# re-define Region_colors so we don't have the additional regions as factors to deal with later
Region_colors = c("#311068","#E95562")

# CLR-transform the matrix for plotting the heatmap
malt_otu_pca_table_clr <- clr(malt_otu_pca_table) # malt_otu_pca_table_filt
malt_otu_pca_table_clr <- as.matrix(malt_otu_pca_table_clr)

malt_otu_pca_table_clr_long <- malt_otu_pca_table_clr %>%
  as.data.frame() %>%
  rownames_to_column("Library_ID") %>%
  gather("Species","CLR",2:ncol(.)) %>%
  spread(Library_ID,CLR) %>%
  filter(Species %in% species_order) %>%
  mutate(Species = fct_relevel(Species, species_order)) %>%
  arrange(Species) %>%
  gather("Library_ID","CLR", 2:ncol(.)) %>%
  inner_join(., full_metadata %>%
               select(Library_ID, Region), by = "Library_ID") %>%
  inner_join(., Region_colors %>%
               as_data_frame(.) %>%
               rename(Color = value) %>%
               mutate(Region = c("Netherlands","Spain")), by = "Region") %>%
  mutate(Library_ID = fct_relevel(Library_ID,  sample_order)) %>% # malt_otu_pca_table_clr_smp_order for cluster order
  arrange(Library_ID)

vector_color = malt_otu_pca_table_clr_long %>%
  select(Library_ID, Color) %>%
  unique %>%
  pull(Color) %>%
  as.character()

# make a heatmap with ggplot
heatmap_base <- ggplot(malt_otu_pca_table_clr_long, aes(Species, Library_ID, fill = CLR)) +
  geom_tile() +
  scale_fill_viridis_c(option = "A") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.95), # , color = vector_color
        plot.margin = margin(0, 0, 0, 0, "pt"),
        axis.title.y = element_blank(),
        legend.position = "left")

heatmap_base


```

Look at the first and last 50 species ordered this way - are there any
physiological characteristics that exlain grouping (like Gram+/-, oxygen
tolerance, etc)?
```{r}
malt_otu.pca$loadings$X %>%
  as.data.frame() %>%
  rownames_to_column("Species") %>%
  arrange(desc(PC1)) %>%
  slice_head(., n = 50)

malt_otu.pca$loadings$X %>%
  as.data.frame() %>%
  rownames_to_column("Species") %>%
  arrange(desc(PC1)) %>%
  slice_tail(., n = 50) %>%
  arrange(PC1)

```
It looks like there are 2 things going on here. First, there is a gradient in
oxygen tolerance, and second there is a difference in preservation. Many of the
first 30 or so taxa with strongest loading in PC1+ values are early-colonizer,
aerobic and facultative taxa. The species with strongest loading in PC1- values
are largely not typical oral taxa, however are anaerobic, assacharolytic, and
close relatives of other (well-known/characterized) oral taxa.Then after that
the taxa are predominantly late-colonizer, anaerobic taxa more commonly seen in
dental plaque.

Kirsten pooled calculus off of multiple teeth for extraction of MID samples, but
I need to ask if Lena has records of which teeth she pooled for each sample.
Also need to ask Lena which teeth were sampled for the CMB individuals, and if
they were pooled.

It would be good to be able to assess preservation. Perhaps run sourceTracker?
The ARS bones from James' paper were included in decontam as controls, but these
bones are not site-specific. Does that matter?

```{r}
pc1_pws_species <- malt_otu.pca$loadings$X %>%
         as.data.frame() %>%
         rownames_to_column("Species") %>%
         arrange(desc(PC1)) %>%
         slice_head(., n = 50) %>%
         pull(Species)

all_species_decontam %>%
  adorn_totals(where = "col") %>%
  select(Species, Total) %>%
  arrange(desc(Total)) %>%
  mutate(Order = 1:n()) %>%
  filter(Species %in% pc1_pws_species) %>%
  as_tibble()

pc1_neg_species <- malt_otu.pca$loadings$X %>%
         as.data.frame() %>%
         rownames_to_column("Species") %>%
         arrange(desc(PC1)) %>%
         slice_tail(., n = 50) %>%
         pull(Species)

all_species_decontam %>%
  adorn_totals(where = "col") %>%
  select(Species, Total) %>%
  arrange(desc(Total)) %>%
  mutate(Order = 1:n()) %>%
  filter(Species %in% pc1_neg_species) %>%
  as_tibble()

```

## Abundance-filtered data
```{r}
malt_otu_filt_pca_table <- all_species_decontam %>%
  select(matches("Species|MID")) %>%
  select(-one_of(outliers)) %>%
  select(-one_of(poor_samples)) %>%
  adorn_totals(where = "col", name = "Sum_species") %>%
  mutate(Percent = Sum_species/sum(Sum_species)*100) %>%
  filter(Percent > 0.001) %>%
  gather("Library_ID","Counts",2:ncol(.)) %>%
  mutate(Counts = Counts + 1) %>%
  spread(Species, Counts) %>%
  column_to_rownames("Library_ID")

# look at the screenplot to decide how many components to keep for the PCA -> will keep all but 2
mixOmics::tune.pca(malt_otu_filt_pca_table, logratio = 'CLR')
# perform a PCA to see how the data cluster
malt_otu_filt.pca <- mixOmics::pca(malt_otu_filt_pca_table, ncomp = 2, logratio = 'CLR')
malt_filt_pca_values <- malt_otu_filt.pca$variates$X %>%
  as.data.frame() %>%
  rownames_to_column("Library_ID") %>%
  inner_join(., full_metadata, by = "Library_ID")

```
There is a point, point where bars jump from tall to much shorter, this number is used to determine numbers of components for pca (2 bars high, dropdown = leave at 2, only one bar = leave at 2), these contain most contaminations. Will leave 2 as number.

Plot the PCA
```{r}
# plot the PCA with Library_IDs
ggplot(malt_filt_pca_values, aes(PC1, PC2, color = Region, shape = Region)) +
  geom_text(aes(label = Library_ID), size = 3) +
  theme_minimal()

horseshoe <- ggplot(malt_filt_pca_values, aes(PC2, PC1, color = Region, shape = Region)) +
  geom_point(size = 2) +
  theme_minimal()

horseshoe

ggsave("~/Desktop/MID_horseshoe.pdf", plot = horseshoe, device = "pdf",
       scale = 1, width = 5, height = 5, units = c("in"), dpi = 300)

```



```{r}

# plot with the species ordered by loading in PC1
# Plot the heatmap with samples ordered by the loading in PC1 from above
# make the Library_IDs a factor so they're in the correct order for the heatmap using the vector from the step above


sample_order <- malt_filt_pca_values %>%
  arrange(PC1) %>%
  pull(Library_ID)

species_filt_order <- malt_otu_filt.pca$loadings$X %>%
  as.data.frame() %>%
  rownames_to_column("Species") %>%
  arrange(desc(PC1)) %>%
  filter(!between(PC1, -0.01, 0.01)) %>%
  pull(Species)

# re-define Region_colors so we don't have the additional regions as factors to deal with later
Region_colors = c("#311068","#E95562")

# CLR-transform the matrix for plotting the heatmap
malt_otu_filt_pca_table_clr <- clr(malt_otu_filt_pca_table) # malt_otu_filt_pca_table_filt
malt_otu_filt_pca_table_clr <- as.matrix(malt_otu_filt_pca_table_clr)

malt_otu_filt_pca_table_clr_long <- malt_otu_filt_pca_table_clr %>%
  as.data.frame() %>%
  rownames_to_column("Library_ID") %>%
  gather("Species","CLR",2:ncol(.)) %>%
  spread(Library_ID,CLR) %>%
  filter(Species %in% species_filt_order) %>%
  mutate(Species = fct_relevel(Species, species_filt_order)) %>%
  arrange(Species) %>%
  gather("Library_ID","CLR", 2:ncol(.)) %>%
  inner_join(., full_metadata %>%
               select(Library_ID, Region), by = "Library_ID") %>%
  inner_join(., Region_colors %>%
               as_data_frame(.) %>%
               rename(Color = value) %>%
               mutate(Region = c("Netherlands","Spain")), by = "Region") %>%
  mutate(Library_ID = fct_relevel(Library_ID,  sample_order)) %>% # malt_otu_filt_pca_table_clr_smp_order for cluster order
  arrange(Library_ID)

vector_color = malt_otu_filt_pca_table_clr_long %>%
  select(Library_ID, Color) %>%
  unique %>%
  pull(Color) %>%
  as.character()

# make a heatmap with ggplot
heatmap_base <- ggplot(malt_otu_filt_pca_table_clr_long, aes(Species, Library_ID, fill = CLR)) +
  geom_tile() +
  scale_fill_viridis_c(option = "A") +
  theme_minimal(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.95), # , color = vector_color
        plot.margin = margin(0, 0, 0, 0, "pt"),
        axis.title.y = element_blank(),
        legend.position = "left")

heatmap_base

ggsave("~/Desktop/heatmap_ordered.pdf", plot = heatmap_base, device = "pdf",
       scale = 1, width = 12, height = 8, units = c("in"), dpi = 300)

```

Look at the first and last 50 species ordered this way - are there any
physiological characteristics that exlain grouping (like Gram+/-, oxygen
tolerance, etc)?
```{r}
malt_otu_filt.pca$loadings$X %>%
  as.data.frame() %>%
  rownames_to_column("Species") %>%
  arrange(desc(PC1)) %>%
  slice_head(., n = 50)

malt_otu_filt.pca$loadings$X %>%
  as.data.frame() %>%
  rownames_to_column("Species") %>%
  arrange(desc(PC1)) %>%
  slice_tail(., n = 50) %>%
  arrange(PC1)

```

```{r}
pc1_pws_species_filt <- malt_otu_filt.pca$loadings$X %>%
         as.data.frame() %>%
         rownames_to_column("Species") %>%
         arrange(desc(PC1)) %>%
         slice_head(., n = 50) %>%
         pull(Species)

all_species_decontam %>%
  adorn_totals(where = "col") %>%
  select(Species, Total) %>%
  arrange(desc(Total)) %>%
  mutate(Order = 1:n()) %>%
  filter(Species %in% pc1_pws_species_filt) %>%
  as_tibble()

pc1_neg_species_filt <- malt_otu_filt.pca$loadings$X %>%
         as.data.frame() %>%
         rownames_to_column("Species") %>%
         arrange(desc(PC1)) %>%
         slice_tail(., n = 50) %>%
         pull(Species)

all_species_decontam %>%
  adorn_totals(where = "col") %>%
  select(Species, Total) %>%
  arrange(desc(Total)) %>%
  mutate(Order = 1:n()) %>%
  filter(Species %in% pc1_neg_species_filt) %>%
  as_tibble()

```

```{r}

malt_pca_species %>%
  mutate(topPC1 = PC1 > PC2 & PC1 > PC3,
         lowPC1 = PC1 < PC2 & PC1 < PC3) %>%
  filter(topPC1 == TRUE, lowPC1 == TRUE)

```






































